Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motions Characteristic to Vertical Axis Turbines

### 2016-06-13 paper.tex
General cleanup; resolves #1

\begin{abstract}
    Vertical axis turbines (VATs) can be used to extract renewable energy from
    wind or water flows. A simpler design, low cost of maintenance, and the
    ability to accept flow from all directions perpendicular to the rotor axis
    are some of the most important advantages over conventional horizontal axis
    turbines (HATs). However, VATs encounter complex and unsteady fluid
    dynamics, which present significant modeling challenges. One of the most
    relevant phenomena is dynamic stall, which is caused by the unsteady
    variation of angle of attack throughout the blade rotation, and is the focus
    of the present study. Dynamic stall is usually used as a passive control for
    VAT operating conditions, hence the importance of predicting its effects. In
    this study, a coupled model is implemented with the open-source CFD
    toolbox OpenFOAM for solving the Navier-Stokes equations, where an actuator
    line model and dynamic stall model are used to compute the blade loading and
    body force. Force coefficients obtained from the model are validated with
    experimental data of pitching airfoil in similar operating conditions as an
    H-rotor type VAT. Numerical results show reasonable agreement with
    experimental data for pitching motion.
\end{abstract}

@@

\section{Methodology}
For solving the governing equations of the phenomena involved, a coupled model have
been implemented: the actuator line model is used to calculate the flow
velocities and thereby the angle of attack. The dynamic stall model is used to
calculate blade force coefficients, which the actuator line model needs for the
velocity field calculations.

@@

\subsection{Actuator Line Model (ALM)}
The ALM is a three-dimensional and unsteady aerodynamic
model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
study the flow around wind turbines. It is a combination of a solver of
Navier-Stokes equations with a so-called actuator line technique, in which
blades of the turbine are represented by a radial distribution of body forces
along lines. These forces are determined using a dynamic stall model commonly
based on empirical data. This work uses the library turbinesFoam developed by
Bachant\cite{bachant2015simulating}. This model is based on incompressible Navier-Stokes equation
\begin{align}
 & \nonumber \\ & \dfrac{\partial \vec{V}}{\partial t} + \vec{V} \cdot \nabla \vec{V} = - \dfrac{1}{ \rho} \nabla p + \nu \nabla ^2 \vec{V} + \vec{f}	  \label{eqmom}
\end{align} %\begin{align}
with $\vec{f}$ representing the body forces, which are the loads on the rotating blades for this case.

Static lift and drag coefficients of the 2D profile used for this study are already known from the technical technical report made by Sheldahl and Klimas \cite{sheldahl1981aerodynamic}, which is a popular database set containing values for a wide range of Reynolds numbers. These coefficients are tabulated and combined with the blade element approach for determining
body forces acting on blades. Relative flow velocity $\vec{V}_{rel} $ and angle of attack $\alpha$ are calculated for each blade using the geometric relation between the velocity of the blade $- \Omega r$, with $\omega$ representing the angular (rotational) velocity of the rotor and $r$ is the radius of the blade element, and the velocity of the local incoming wind flow $\vec{V}_{in}$. Figure \ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the plane perpendicular to the turbine axis.

@@

Once the angle of attack, relative velocity are obtained, the blade element lift and drag forces per spanwise length unit are calculated as
\begin{align}
& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{rel} \right|^2  \label{lift}
\end{align} %\begin{align}
\begin{align}
& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{rel} \right|^2  \label{drag}
\end{align} %\begin{align}
where $C_L$ and $C_D$ are the lift and drag coefficients, respectively. Both are function of the Reynolds number and the angle of attack. It should be emphasized that the direction of the lift and drag are respectively perpendicular and parallel to the relative velocity $\vec{V}_{rel}$. $c$ represents the chord length.

Same procedure is used to calculate forces from the turbine shaft and blade support struts. After all these forces on the actuator lines are calculated, then are added as a body force source, per unit of density (incompressible assumption), in the momentum Navier-Stokes equations (equation \ref{eqmom}).

\subsection{Dynamic Stall Modeling (DSM)} The DSM is represented by the Leishman-Beddoes
model\cite{leishman1986generalised} with the modifications of Sheng et
al\cite{sheng2008modified} and Dyachuk\cite{dyachuk}.
It is capable to calculate the unsteady lift,
pitching moment and drag, giving a physical description of the aerodynamics. It
have been validated with experimental data in\cite{leishman1989semi}.
The model is separated into three subsystems: an attached flow model for unsteady linear
airloads, a separated flow model for non-linear airloads and a dynamic stall
model for the airloads induced by the leading edge vortex.

@@

\section{Conclusions}
It have been demonstrated that the presented coupled model ALM with DSM, shows good agreement with experimental data, considering the complexity
of the flows studied. Moreover, the model have been tested for a wide range of pithing amplitudes, therefore, with different stall conditions (not, small and deep stall), keeping a good results accuracy compared with experiments.
The model is stable and accurate for the teste pithing blade motions.
This makes the model suitable for application in VAWT
simulations.

>>>>

\begin{abstract}
    Vertical axis turbines (VATs) can be used to extract renewable energy from
    wind or water flows. A simpler design, low cost of maintenance, and the
    ability to accept flow from all directions perpendicular to the rotor axis
    are some of the most important advantages over conventional horizontal axis
    turbines (HATs). However, VATs encounter complex and unsteady fluid
    dynamics, which present significant modeling challenges. One of the most
    relevant phenomena is dynamic stall, which is caused by the unsteady
    variation of angle of attack throughout the blade rotation, and is the focus
    of the present study. Dynamic stall is usually used as a passive control for
    VAT operating conditions, hence the importance of predicting its effects. In
    this study, a coupled model is implemented with the open-source CFD
    toolbox OpenFOAM for solving the Navier--Stokes equations, where an actuator
    line model and dynamic stall model are used to compute the blade loading and
    body force. Force coefficients obtained from the model are validated with
    experimental data of pitching airfoil in similar operating conditions as an
    H-rotor type VAT. Numerical results show reasonable agreement with
    experimental data for pitching motion.
\end{abstract}

@@

\section{Methodology}

For solving the governing equations of the phenomena involved, a coupled model
has been implemented: the actuator line model (ALM) samples the flow velocity
from the Navier--Stokes solver, and thereby computes each blade element's angle
of attack and relative velocity. The dynamic stall model is used to calculate
blade force coefficients, which the actuator line model uses to impart the body
force back into the flow solver.

@@

\subsection{Actuator Line Model (ALM)}

The ALM is a three-dimensional and unsteady aerodynamic model developed by
S{\o}rensen and Shen\cite{sorensen1999computation}, used to study the flow
around wind turbines. It is a combination of a solver of the Navier--Stokes equations
with a so-called actuator line technique, in which blades of the turbine are
represented by a distribution of body forces along lines. These forces
are determined using a dynamic stall model commonly based on empirical data.
This work uses the library turbinesFoam developed by Bachant et
al.~\cite{bachant2015simulating,Bachant2016-VAT-ALM}.

The model employed here is based on the incompressible Navier--Stokes equation
\begin{align}
 & \nonumber \\ & \dfrac{\partial \vec{V}}{\partial t} + \vec{V} \cdot \nabla \vec{V} = - \dfrac{1}{ \rho} \nabla p + \nu \nabla ^2 \vec{V} + \vec{f}	  \label{eqmom}
\end{align} %\begin{align}
with $\vec{f}$ representing the body forces, which are the loads on the rotating blades for this case.

Static lift and drag coefficients of the 2D profile used for this study are
taken from the technical technical report made by Sheldahl and Klimas
\cite{sheldahl1981aerodynamic}, which is a popular database containing values
for a wide range of Reynolds numbers. These coefficients are tabulated and
combined with the blade element approach for determining body forces acting on
blades. Relative flow velocity $\vec{V}_{\mathrm{rel}}$ and angle of attack
$\alpha$ are calculated for each blade using the geometric relation between the
velocity of the blade $- \omega r$, with $\omega$ representing the angular
(rotational) velocity of the rotor and $r$ the radius of the blade element, and
the velocity of the local incoming wind flow $\vec{V}_{\mathrm{in}}$. Figure
\ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the
plane perpendicular to the turbine axis.

@@

Once the angle of attack, relative velocity are obtained, the blade element lift and drag forces per spanwise length unit are calculated as
\begin{align}
& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{rel} \right|^2  \label{lift}
\end{align} %\begin{align}
\begin{align}
& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{rel} \right|^2  \label{drag}
\end{align} %\begin{align}
where $C_L$ and $C_D$ are the lift and drag coefficients, respectively. Both are
function of the Reynolds number and the angle of attack. It should be emphasized
that the direction of the lift and drag are respectively perpendicular and
parallel to the relative velocity $\vec{V}_{\mathrm{rel}}$. $c$ represents the
chord length.

Same procedure is used to calculate forces from the turbine shaft and blade
support struts. After all these forces on the actuator lines are calculated,
then are added as a body force source, per unit of density (incompressible
assumption), in the momentum Navier--Stokes equations (equation \ref{eqmom}).

\subsection{Dynamic Stall Modeling (DSM)} The DSM is represented by the
Leishman-Beddoes model~\cite{leishman1986generalised} with the modifications of
Sheng et al.~\cite{sheng2008modified} and Dyachuk~\cite{dyachuk}. It is capable to
calculate the unsteady lift, pitching moment and drag, giving a physical
description of the aerodynamics. It have been validated with experimental data
in~\cite{leishman1989semi}. The model is separated into three subsystems: an
attached flow model for unsteady linear airloads, a separated flow model for
non-linear airloads and a dynamic stall model for the airloads induced by the
leading edge vortex.

@@

\section{Conclusions}

It has been demonstrated that the presented coupled model ALM with DSM, shows
good agreement with experimental data, considering the complexity of the flows
studied. Moreover, the model has been tested for a wide range of pithing
amplitudes, therefore, with different stall conditions (none, shallow, and deep
stall), results agreed well compared with experiments. The model is stable and
accurate for the tested pithing blade motions, which makes it suitable for
application in VAT simulations.
